A Generalized Sphere Theorem for Positively Curved Combinatorial 3-Manifolds

### 2013-04-13 paper/paper.tex
more writing

\begin{thm}[Grove-Shiohama] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with section curvature $K \geq \delta > 0$ and $diam(M) > \frac{\pi}{2\sqrt{\delta}}$. Then, $M$ is homeomorphic to a sphere.
\label{Grove-Shiohama}
\end{thm}

\noindent Note that the diameter bound in this theorem is exactly half that allowed by the Bonnet-Myers theorem:

\begin{thm}[Bonnet-Myers] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with section curvature $K \geq \delta > 0$. Then $diam(M) \leq \frac{\pi}{\sqrt{\delta}}$.
\end{thm}

@@

The discrete version of Theorem \ref{Grove-Shiohama} given in this paper will apply to positively curved combinatorial 3-manifolds. A {\em combinatorial 3-manifold} is an abstract simplicial complex in which the link of each vertex is a 2-sphere. We call such a space {\em positively curved} if at most five tetrahedra are incident along each edge. Why this terminology? If we endow a combinatorial 3-manifold with the standard piecewise-linear (PL) metric (in which all edges have unit-length), this condition is equivalent to requiring an angle deficit along each edge. In classical differential geometry an angle deficit is intimately related to positive curvature.

@@

\begin{dfn}The {\em edge-distance} between two vertices $v$ and $w$ in an abstract simplicial complex $M$ is the minimum number of edges needed in an edge path from $v$ to $w$. We denote this by $d_1(v,w)$. The {\em edge-diameter} of $M$, written as $diam_1(M)$, is the maximum of $d_1(v,w)$ over all pairs of vertices $v$ and $w$ in $M$. 
\end{dfn}

\noindent Previous work \cite{Trout10} has already demonstrated a discrete version of the Bonnet-Myers theorem which applies to positively curved combinatorial 3-manifolds and which uses these notions of distance and diameter.

\begin{thm}[Trout] A positively curved combinatorial 3-manifold $M$ has edge-diameter at most five.
\label{discrete_BM}
\end{thm}

While Theorem \ref{discrete_BM} is stated in terms of edge-diameter, the proof is greatly facilitated by expanding the set of paths under consideration to include paths containing not only edges, but PL-paths between vertices called {\em hops} and {\em jumps}.

\begin{dfn}[Jumps] Suppose there are edges $e_1$ and $e_2$ and vertices $v_1$ and $v_2$ in $M$ so that $e_1*e_2$ is a 3-simplex in $M$ and $v_1*e_1$ and $v_2*e_2$ are 2-simplices in $M$. We call the PL-path from $v_1$ through the barycenters of $e_1$ and $e_2$ and ending on $v_2$ a {\em jump} from $v_1$ to $v_2$. See Figure \ref{jump_fig}.
\end{dfn}

\begin{dfn}[Hops] Suppose $\tau$ is a 2-simplex in $M$ and $v_1$ and $v_2$ vertices in $M$ such that $v_1*\tau$ and $v_2*\tau$ are 3-simplices in $M$. The PL-path from $v_1$ through the barycenter of $\tau$ and ending on $v_2$ will be called a {\em hop} from $v_1$ to $v_2$. See Figure \ref{hop_fig}.
\end{dfn}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.6\linewidth]{figures/hops.pdf}
        \caption{A {\em hop} from vertex $v_1$ to vertex $v_2$, transverse to the 2-simplex $\tau$ TO DO: CROP OUT LEFT IMAGE}
    \end{center}
    \label{hop_fig}
\end{figure}

\begin{figure}
    \begin{center}
        \includegraphics[width=0.4\linewidth]{figures/jump.pdf}
        \caption{A {\em jump} from vertex $v_1$ to vertex $v_2$, transverse to the edges $e_1$ and $e_2$}
    \end{center}
    \label{jump_fig}
\end{figure}
\section{Results}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.6\linewidth]{figures/global_statistics.png}
    \caption{Global statistics}
    \end{center}
\end{figure}

>>>>

\begin{thm}[Grove-Shiohama] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with section curvature $K \geq \delta > 0$ and diameter greater than $\frac{\pi}{2\sqrt{\delta}}$. Then, $M$ is homeomorphic to a sphere.
\label{thm:grove_shiohama}
\end{thm}

\noindent Note that this bound is sharp, since the real projective space $\mathbb{RP}^n$ admits a metric with uniform sectional curvature $K=1$ and diameter $\pi/2$. We should also note also that the diameter bound in Theorem \ref{thm:gove_shiohama} is exactly half the maximum diameter allowed by the Bonnet-Myers theorem:

\begin{thm}[Bonnet-Myers] Let $M$ be a complete, connected, $n$-dimensional Riemannian manifold with section curvature $K \geq \delta > 0$. Then the diameter of $M$ is at most $\frac{\pi}{\sqrt{\delta}}$.
\label{thm:bonnet_myers}
\end{thm}

@@

The discrete version of Theorem \ref{thm:grove_shiohama} given in this paper will apply to positively curved combinatorial 3-manifolds. A {\em combinatorial 3-manifold} is an abstract simplicial complex in which the link of each vertex is a 2-sphere. We call such a space {\em positively curved} if at most five tetrahedra are incident along each edge. Why this terminology? If we endow a combinatorial 3-manifold with the standard piecewise-linear (PL) metric (in which all edges have unit-length), this condition is equivalent to requiring an angle deficit along each edge. In classical differential geometry an angle deficit is intimately related to positive curvature.

@@

\begin{dfn}The {\em edge-distance} between two vertices $v$ and $w$ in an abstract simplicial complex $M$ is the minimum length, in edges, of an edge path from $v$ to $w$. We denote this quantity $d_1(v,w)$. The {\em edge-diameter} of $M$, written as $diam_1(M)$, is the maximum of $d_1(v,w)$ over all pairs of vertices $v$ and $w$ in $M$. 
\end{dfn}

\noindent Previous work \cite{Trout10} has demonstrated a discrete version of the Bonnet-Myers theorem which applies to positively curved combinatorial 3-manifolds and which uses these notions of distance and diameter.

\begin{thm}[Trout] For any positively curved combinatorial 3-manifold $M$ we have $diam_1(M)\leq 5$. Moreover, this bound is sharp.
\label{thm:discrete_BM}
\end{thm}

While Theorem \ref{thm:discrete_BM} is stated in terms of edge-diameter, the proof is greatly facilitated by expanding the set of paths under consideration to include those which contain not only edges, but also other types of PL-paths between vertices. The first new type of path is called a {\em hop}.

\begin{dfn}[Hops] Suppose $\tau$ is a 2-simplex in $M$ and $v_1$ and $v_2$ vertices in $M$ such that $v_1*\tau$ and $v_2*\tau$ are 3-simplices in $M$. The PL-path from $v_1$ through the barycenter of $\tau$ and ending on $v_2$ will be called a {\em hop} from $v_1$ to $v_2$. See Figure \ref{hop}. TO DO: WHY DOES THIS SAY FIGURE 2?!?!
\end{dfn}

\noindent The other new type of path we call a {\em jump.}

\begin{dfn}[Jumps] Suppose there are edges $e_1$ and $e_2$ and vertices $v_1$ and $v_2$ in $M$ so that $e_1*e_2$ is a 3-simplex in $M$ and $v_1*e_1$ and $v_2*e_2$ are 2-simplices in $M$. We call the PL-path from $v_1$ through the barycenters of $e_1$ and $e_2$ and ending on $v_2$ a {\em jump} from $v_1$ to $v_2$. See Figure \ref{jump}.
\end{dfn}

\noindent The length of each hop and edge will be its length as a PL-path in the standard PL metric. Some basic Euclidean geometry tells us these lengths are $H = \frac{2}{3}\sqrt{2}$ and $J = \frac{1}{2}\sqrt{2} + \sqrt{3}$ respectively. We let $d(v,w)$ denote the distance between vertices $v$ and $w$ obtained by minimizing over all paths containing edges as well as hops and jumps. Similarly, we let $diam(M)$ denote the diameter defined in terms of this new distance.

Why are these paths relevant to this paper? The classical generalized sphere theorem requires the manifold to have diameter more than half the maximum allowed by the Bonnet-Myers theorem. If we were to naively imitate the discrete Bonnet-Myers result, Theorem \ref{thm:discrete_BM}, the corresponding diameter inequality would be $diam_1(M)>2.5$. Since only integral values of edge-distance can occur, this would correspond to $diam(M)\geq 3$. Unfortunately, this bound will not work since there are triangulations of $\mathbb{RP}^3$ with edge-diameter three. However, if we instead use the finer-grained measure of diameter, $diam(M)$ this problem disappears.

\begin{figure}
	\label{hop}
    \begin{center}
        \includegraphics[width=0.6\linewidth]{figures/hops.pdf}
        \caption{A {\em hop} from vertex $v_1$ to vertex $v_2$. TO DO: CROP OUT LEFT IMAGE}
    \end{center}
\end{figure}

\begin{figure}
    \label{jump}
    \begin{center}
        \includegraphics[width=0.4\linewidth]{figures/jump.pdf}
        \caption{A {\em jump} from vertex $v_1$ to vertex $v_2$}
    \end{center}
\end{figure}

\section{Results}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.6\linewidth]{figures/global_statistics.png}
    \caption{Global statistics}
    \end{center}
    \label{global_statistics}
\end{figure}
